Predicting the 2024 US Presidential Election with a Model-Based Forecast

### 2024-11-04 paper/paper.qmd
finalize paper

---
title: "Predicting the 2024 US Presidential Election with a Model-Based Forecast"
subtitle: "Kamala Harris Predicted to Lead with 341 Electoral Votes Based on a Poll-Based Prediction"
author: 
  - Yuanyi (Leo) Liu
  - Dezhen Chen
  - Ziyuan Shen
thanks: "Code and data are available at: [Forecasting the 2024 US Presidential Election](https://github.com/leoyliu/Forecasting-the-2024-US-Presidential-Election)."
date: today
date-format: long
abstract: "In this paper, we develop a linear model to predict the outcome of the 2024 U.S. presidential election using a \"poll-of-polls\" approach that aggregates data from multiple well-known pollsters. By analyzing state-level polling trends, our model predicts that Kamala Harris is likely to secure key states, leading to an estimated electoral vote count of 341 for Harris and 197 for Donald Trump. This approach, which combines multiple polls, reduces bias and improves reliability, providing a more accurate picture of public sentiment leading up to the election."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Our findings suggest that while Trump has a competitive chance, Harris is projected to win key states and receive a majority of electoral votes. Harris is estimated to receive 341 electoral votes to Trump's 197, and the uncertainty around these estimates is represented by confidence intervals, which take into account polling changes and model assumptions. This is important because accurate election forecasts help reduce uncertainty, which allows political stakeholders to allocate resources efficiently and gives the public a better understanding of likely outcomes. Aggregate polling reduces uncertainty and provides a clearer picture of potential outcomes.

@@

We use the statistical programming language R [@citeR] to process and analyze polling data for the 2024 U.S. Presidential election. The dataset used for this analysis was obtained from the FiveThirtyEight 2024 U.S. Presidential Election Polls [@citedata]. It consists of polling data for the 2024 general election, covering various polling organizations and methodologies. Following methodologies discussed by "Telling Stories with Data" [@tellingstories], we forecast election outcomes using the "poll of polls" method, which aggregates results from multiple polls to reduce bias and provide a more accurate representation of voter sentiment. For key operations, please refer to [Appendix -@sec-data-cleaning].

@@

ggplot(cleaned_data, aes(x = end_date, y = pct)) +
  geom_point(alpha = 0.5) +  # Add points for individual poll data
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +  # Show month and year, with one label per month
  labs(
    x = "Date of Poll",
    y = "Polling Percentage (%)",
  ) +
  theme_minimal() +  # Use a clean minimal theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),  # Center and adjust title size
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  ) +
  geom_smooth()
```

Using package `ggplot2` [@citeggplot], @fig-trump-pct shows Donald Trump's polling percentages over time from July to October 2024. Each black dot represents an individual poll conducted on a specific date, indicating the percentage of voters who supported Trump in that poll. The blue line represents a trendline, smoothing out the individual poll results to show the general trajectory of Trump's support over time.

Initially, Trump's polling percentage appears stable with small fluctuations around 45%. However, as the campaign progresses into October, there is a slight upward trend in Trump's polling percentage, suggesting that his support increased as the election neared.

@@

# Create a bar chart for the distribution of numeric_grade
p1 <- ggplot(cleaned_data, aes(x = numeric_grade)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(x = "Numeric Grade",
       y = "Count") +
  theme_minimal()

# Create a bar chart for the distribution of transparency_score
p2 <- ggplot(cleaned_data, aes(x = transparency_score)) +
  geom_bar(fill = "darkorange", color = "black") +
  labs(x = "Transparency Score",
       y = "Count") +
  theme_minimal()

@@

By using package `usmap` [@citeusmap], @fig-statemap displays the predicted outcome for each state in the 2024 U.S. Presidential Election, with red representing states where Trump is expected to win and blue for those favoring Harris.

It shows Trump gaining support across central and southern states, including Texas and Missouri, which have traditionally leaned Republican. Harris, meanwhile, is projected to lead in states along the West Coast and in the Northeast, such as California and New York, regions with a history of Democratic support. This visual representation complements the previous electoral vote projection, where Harris is anticipated to surpass the 270-vote threshold needed to win the presidency.

>>>>

---
title: "Predicting the 2024 US Presidential Election with a Model-Based Forecast"
subtitle: "Kamala Harris Predicted to Lead with 341 Electoral Votes Based on a Poll-Based Prediction"
author: 
  - Yuanyi (Leo) Liu
  - Dezhen Chen
  - Ziyuan Shen
thanks: "Code and data are available at: [Forecasting the 2024 US Presidential Election](https://github.com/leoyliu/Forecasting-the-2024-US-Presidential-Election)."
date: today
date-format: long
abstract: "In this paper, we develop a linear model to predict the outcome of the 2024 U.S. presidential election using a \"poll-of-polls\" approach that aggregates data from multiple pollsters. By analyzing state-level polling trends, our model predicts that Kamala Harris is likely to secure key states, leading to an estimated electoral vote count of 341 for Harris and 197 for Donald Trump. This approach, which combines multiple polls, reduces bias and improves reliability, providing a more accurate picture of public sentiment leading up to the election."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Our findings suggest that while Trump has a competitive chance, Harris is projected to win key states and receive a majority of electoral votes. Harris is estimated to receive 341 electoral votes to Trump's 197, and the uncertainty around these estimates is represented by confidence intervals, which take into account polling changes and model assumptions. This is important because accurate election forecasts help reduce uncertainty, which allows political stakeholders to allocate resources efficiently and gives the public a better understanding of likely outcomes.

@@

We use the statistical programming language R [@citeR] to process and analyze polling data for the 2024 U.S. Presidential election. The dataset used for this analysis was obtained from the FiveThirtyEight 2024 U.S. Presidential Election Polls [@citedata] as of October 30, 2024. It consists of polling data for the 2024 general election, covering various polling organizations and methodologies. Following methodologies discussed by "Telling Stories with Data" [@tellingstories], we forecast election outcomes using the "poll of polls" method, which aggregates results from multiple polls to reduce bias and provide a more accurate representation of voter sentiment. For key operations, please refer to [Appendix -@sec-data-cleaning].

@@

ggplot(cleaned_data, aes(x = end_date, y = pct)) +
  geom_point(alpha = 0.5, color = "black", size = 1.5) + 
  geom_smooth(color = "#FF6F61", fill = "#FF6F61", method = "loess", se = TRUE, linetype = "solid", size = 1.2) + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y", expand = c(0.01, 0.01)) + 
  labs(
    x = "Date of Poll",
    y = "Polling Percentage (%)",
  ) +
  theme_minimal(base_size = 10) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    axis.text = element_text(color = "gray20"),  # Darker axis text for contrast
    panel.grid.minor = element_blank()  # Remove minor grid for a cleaner look
)
```

Using package `ggplot2` [@citeggplot], @fig-trump-pct shows Donald Trump's polling percentages over time from July to October 2024. Each black dot represents an individual poll conducted on a specific date, indicating the percentage of voters who supported Trump in that poll. Initially, Trump's polling percentage appears stable with small fluctuations around 45%. However, as the campaign progresses into October, there is a slight upward trend in Trump's polling percentage, suggesting that his support increased as the election neared.

@@

# Create a bar chart for the distribution of numeric_grade
p1 <- ggplot(cleaned_data, aes(x = numeric_grade)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(x = "Numeric Grade",
       y = "Frequency") +
  theme_minimal()

# Create a bar chart for the distribution of transparency_score
p2 <- ggplot(cleaned_data, aes(x = transparency_score)) +
  geom_bar(fill = "darkorange", color = "black") +
  labs(x = "Transparency Score",
       y = "Frequency") +
  theme_minimal()

@@

By using package `usmap` [@citeusmap], @fig-statemap displays the predicted outcome for each state in the 2024 U.S. Presidential Election. It shows Trump gaining support across central and southern states, including Texas and Missouri, which have traditionally leaned Republican. Harris, meanwhile, is projected to lead in states along the West Coast and in the Northeast, such as California and New York, regions with a history of Democratic support. This visual representation complements the previous electoral vote projection, where Harris is anticipated to surpass the 270-vote threshold needed to win the presidency.
